---
title: Mixed Effect
output: html_document
---

```{R setup, include = FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE)
```

# Linear Mixed-Effects Models

### A Simple Example of Random Effects

```{r}
library(nlme)
data("Rail")
plot( Rail )
Rail
```
singl-mean model
$$y_{ij} = \beta + \epsilon_{ij},\quad i=1,\dots,M,\quad j=1,\dots,n_i.$$
```{r}
fm1Rail.lm <- lm( travel ~ 1, data = Rail )
fm1Rail.lm
boxplot(fm1Rail.lm$residuals ~ Rail$Rail)
```
fixed-effects model
$$y_{ij} = \beta_i + \epsilon_{ij},\quad i=1,\dots,M,\quad j=1,\dots,n_i.$$
```{r}
fm2Rail.lm <- lm( travel ~ Rail - 1, data = Rail )
fm2Rail.lm
boxplot(fm2Rail.lm$residuals ~ Rail$Rail)
```
random-effects model
$$\begin{aligned}
y_{ij} &= \beta + b_i + \epsilon_{ij},\quad i=1,\dots,M,\quad j=1,\dots,n_i,\\
b_i &\sim N(0,\sigma_b^2),\quad \epsilon_{ij}\sim N(0,\sigma^2)
\end{aligned}$$
```{r}
fm1Rail.lme <- lme( travel ~ 1, data = Rail, random = ~ 1 | Rail )
summary(fm1Rail.lme)
```

```{r}
fm1Rail.lmeML <- update( fm1Rail.lme, method = "ML" )
summary(fm1Rail.lmeML)
```

```{r}
plot(fm1Rail.lme)
intervals(fm1Rail.lme)
anova(fm1Rail.lme)
```

###  A Randomized Block Design
```{r}
data("ergoStool")
plot( ergoStool )
plot.design(ergoStool)
```
fixed-effects model

$$\begin{aligned}
y_{ij} & = \beta_j + b_i + \epsilon_{ij},\quad i=1,\dots,M,\quad j=1,\dots,n_i,\\
b_i &\sim N(0,\sigma_b^2),\quad \epsilon_{ij}\sim N(0,\sigma^2)
\end{aligned}$$

helmert contrasts
```{r}
options( contrasts = c( factor = "contr.helmert", ordered = "contr.poly" ) )
contrasts( ergoStool$Type )
fm1Stool <- lme(effort ~ Type, data = ergoStool, random =~1| Subject)
summary( fm1Stool )
anova( fm1Stool )
```
treatment contrasts
```{r}
options( contrasts = c( factor = "contr.treatment", ordered = "contr.poly" ) )
contrasts( ergoStool$Type )
fm2Stool <- lme(effort ~ Type, data = ergoStool, random =~1| Subject)
summary( fm2Stool )
anova( fm2Stool )
```
$$H_0:\quad \beta_2=\beta_3=\beta_4=0$$
cell-mean model
```{r}
fm3Stool <- lme(effort ~ Type - 1, data = ergoStool, random = ~ 1 | Subject)
summary( fm3Stool )
anova( fm3Stool )
```
$$H_0:\quad \beta_1 = \beta_2=\beta_3=\beta_4=0$$

- The overall effect of the factor should be assessed with anova, which does not depend on the choice of contrasts as long as the intercept term is retained in the model.
- Interpretation of the parameter estimates for a fixed-effects term depends on the contrasts being used.
- For REML estimation, likelihood-ratio tests or comparisons of AIC
or BIC require the same fixed-effects structure and the same choice
of contrasts in all models.

```{r}
intervals( fm1Stool )
plot( fm1Stool, form = resid(., type = "p") ~ fitted(.) | Subject, abline = 0 )
```

### Mixed-Effects Models for Replicated, Blocked Designs
```{r}
data("Machines")
plot( Machines )
attach( Machines ) 
interaction.plot( Machine, Worker, score, las = 1) 
detach()
```

$$\begin{aligned}
y_{ijk} & = \beta_j + b_i + b_{ij} + \epsilon_{ijk},\\
b_i &\sim N(0,\sigma_b^2),\quad \epsilon_{ijk}\sim N(0,\sigma^2)
\end{aligned}$$
```{r}
fm1Machine <- lme( score ~ Machine, data = Machines, random =~1| Worker )
fm1Machine
intervals( fm1Machine )
```


$$\begin{aligned}
y_{ijk} & = \beta_j + b_i + \epsilon_{ijk},\\b_i &\sim N(0,\sigma_1^2),\quad b_{ij} \sim N(0,\sigma_2^2),\quad \epsilon_{ijk}\sim N(0,\sigma^2)
\end{aligned}$$
```{r}
fm2Machine <- update( fm1Machine, random =~1| Worker/Machine )
fm2Machine
intervals( fm2Machine )
```
```{r}
anova( fm1Machine, fm2Machine)
```

unbalanced data
```{r}
## delete selected rows from the Machines data
MachinesUnbal <- Machines[ -c(2,3,6,8,9,12,19,20,27,33), ]
## check that the result is indeed unbalanced
table(MachinesUnbal$Machine, MachinesUnbal$Worker)
fm1MachinesU <- lme( score ~ Machine, data = MachinesUnbal,
                     random =~1| Worker/Machine )
fm1MachinesU
intervals( fm1MachinesU )
```

random interation effects
$$\begin{aligned}
y_{i} & = \mathbf{X}_i\mathbf{\beta} + \mathbf{Z}_i\mathbf{b}_i +  \mathbf{\epsilon}_{i},\\
\mathbf{b}_i &\sim N(0,\Psi),\quad \mathbf{\epsilon}_{i}\sim N(0,\sigma^2\mathbf{I})
\end{aligned}$$
```{r}
fm3Machine <- update( fm1Machine, random = ~Machine - 1 |Worker)
summary( fm3Machine )
```

```{r}
anova( fm1Machine, fm2Machine, fm3Machine )
```

### An Analysis of Covariance Model
$$\begin{aligned}
y_{ij} & = \beta_1 + b_i + \beta_2x_{ij} + \epsilon_{ij},\\
b_i &\sim N(0,\sigma_b^2),\quad \epsilon_{ij}\sim N(0,\sigma^2)
\end{aligned}$$
```{r}
data("Orthodont")
plot( Orthodont )
names( Orthodont )
levels( Orthodont$Sex )
OrthoFem <- Orthodont[ Orthodont$Sex == "Female", ]
```

```{r}
fm1OrthF.lis <- lmList( distance ~ age, data = OrthoFem )
coef( fm1OrthF.lis )
plot( intervals ( fm1OrthF.lis ) )
```
center the data
```{r}
fm2OrthF.lis <- update( fm1OrthF.lis, distance ~ I( age - 11 ) )
plot( intervals( fm2OrthF.lis ) )
```

mixed-effects model
```{r}
fm1OrthF <- lme( distance ~ age, data = OrthoFem, random =~1| Subject )
summary( fm1OrthF )
```
the REML criterion was created to compensate for the downward bias of the maximum
likelihood estimates of variance components, so it should produce larger
estimates.
```{r}
fm1OrthFM <- update( fm1OrthF, method = "ML" )
summary( fm1OrthFM )
```

random effect for intercept and slope
```{r}
fm2OrthF <- update( fm1OrthF, random = ~ age | Subject )
anova( fm1OrthF, fm2OrthF )
```
best linear unbiased predictions (BLUPs)
```{r}
random.effects( fm1OrthF )
# ranef( fm1OrthF )
coefficients( fm1OrthF )
plot(compareFits(coef(fm1OrthF), coef(fm1OrthFM)))
plot( augPred(fm1OrthF), aspect = "xy", grid = T )
```

### Models for Nested Classification Factors
$$y_{ijk} = \beta_1 + \beta_2 d_{ik} + \beta_3 d^2_{ik} + b_{i,1} + b_{i,2}d_{ik} + b_{ij} + \epsilon_{ijk}$$
```{r}
data("Pixel")
plot( Pixel )
fm1Pixel <- lme( pixel ~ day + day^2, data = Pixel,
                 random = list( Dog = ~ day, Side =~1))
intervals( fm1Pixel )
plot( augPred( fm1Pixel ) )
VarCorr( fm1Pixel )
summary( fm1Pixel )
```
```{r}
fm2Pixel <- lme( pixel ~ day + day^2, data = Pixel,
                 random =  ~ day | Dog )
anova( fm1Pixel, fm2Pixel )
fm3Pixel <- update( fm1Pixel, random =~1| Dog/Side )
anova( fm1Pixel, fm3Pixel )
fm4Pixel <- update( fm1Pixel, pixel ~ day + day^2 + Side )
summary( fm4Pixel )
```

### A Split-Plot Experiment
```{r}
data("Oats")
plot( Oats )
fm1Oats <- lme( yield ~ ordered(nitro) * Variety, data = Oats,
                random =~1| Block/Variety )
anova( fm1Oats )
```
remove interaction
```{r}
fm2Oats <- update( fm1Oats, yield ~ ordered(nitro) + Variety )
anova( fm2Oats )
summary( fm2Oats )
```
remove fixed effect
```{r}
fm3Oats <- update( fm1Oats, yield ~ ordered( nitro ) )
summary( fm3Oats )
```
remove nonlinear
```{r}
fm4Oats <- lme( yield ~ nitro, data = Oats, random = ~ 1 | Block/Variety )
summary( fm4Oats )
VarCorr( fm4Oats )
plot( augPred( fm4Oats ), aspect = 2.5, layout = c(6, 3),
      between = list( x = c(0, 0, 0.5) ) )
```
